An Accurate Model for SARS-CoV-2 pooled RT-PCR Tests Errors

### 2021-03-16 comment.tex
better

\section{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development techniques that of pooling schemes of RT-PCR tests
\cite{DorfmanYuvalDor, PoolSize30, BayesianDorfman, MatrixPooling,
  LionDorfman}. One such popular scheme is Dorfman pooling
\cite{DorfmanOriginal, DorfmanYuvalDor}: Select $N$ individuals and
perform a single RT-PCR test on their combined (\emph{pooled})
samples. If the pooled test yields a positive result --- test each
individual separately. The throughput efficiency of Dorfman pooling
has been demonstrated empirically \cite{DorfmanYuvalDor}.

Pooling (or RT-PCR tests, for that matter) is not perfect, and there
is always some nonzero probability of error. Tyipically, authors
assume the probability of a true-positive (the test's
\emph{sensitivity}) does not depend on the number of infected samples
in the pool, but rather on the existence of at least one such
sample. Thus, the probability of a positive result in a pooled test is
(assumed) identical for a pool with one sample from an infected
individual and (e.g.) five such samples. This assumption is common in
the group testing literature \cite{Kim, OptimalDorfmanPool}, as well
as in more specific, COVID-19 focused studies \cite{Simplistic1,
  Simplistic2}. In this study we refute these commonly made
assumptions and show how they affect estimation of false-negative
rates for Dorfman pooling.

@@

\subsection{Common assumptions}
Previous studies of group testing strategies assumed that the
false-negative probability does not depend on the number of infected
samples, but merely on the existence of at least one such sample in a
pool \cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ similar assumptions
\cite{Simplistic1, Simplistic2}. Specifically, these studies assume
that the probability of a negative result is the same for a pool with
a single sample coming from an infected individual and (e.g.)
five. We refute these assumptions and suggest more plausible ones in
Section~\ref{subsec:ours} below.

\subsection{Our model}\label{subsec:ours}
We assume two pathways for a positive pooled test result: Viral RNA
from an infected individual is correctly detected; or, some erroneous
detection occurs (e.g. contaminant viral RNA is introduced). We assume
a homogeneous and disconnected population (each individual is infected
independently and with equal probability).

Our assumptions, although natural, stand in contrast to assumptions
commonly made in the literature \cite{Kim, Simplistic1, Simplistic2,
  OptimalDorfmanPool}. It is commonly assumed that the probability of
a positive pooled test is not increased by having more than one
infected individual in the pool. We refute the common assumptions with
experimental data summarized in Table \ref{table}, collected from
\cite{Salazar}. There, the authors investigate Dorfman pooling and,
regardless of the pooled test result, follow up and test each pool
member separately. Here, we focus on 128 pools for which at least one
subsequent separate test was positive --- of which 29 pooled tests
were negative and 99 positive. In the data cited in \cite{Salazar}, of
the 29 negative pools, subsequent separate testing yielded a single
positive result in 24. In contrast, of the 99 positive pools, 42
yielded a single positive test upon subsequent separate testing.

The data in table \ref{table} allows us to test the following null
hypothesis $H_0$: The probability of a pooled false-negative is equal
for pools with one subsequent positively tested member and pools with
two or more such members. We apply Fisher's exact test for the
presence of more than one positive individual in correctly identified
pools. Fisher's test yields an increased odds ratio of 6.4, 95\% CI
(2.2,23.4), with a p-value $\approx 10^{-4}$. Thus we reject $H_0$,
refuting the independence of a pooled result from one or more
individuals infected within the pool, as assumed in \cite{Simplistic1,
  Simplistic2, OptimalDorfmanPool, Kim}

@@

Consider a pool containing $N$ individuals $\{1,\dots,N\}$. We denote
the true infection states $\theta \in \{0,1\}^N$, so individual $i$ is
infected iff $\theta_i=1$. Data are denoted $\data$. The tests
false-negative rate is denoted $\fn$ and the false-positve rate is
denoted $\fp$. The probability of a negative pooled test result is
presented in \eqref{eq:likelihood a} (along with its complement
\eqref{eq:likelihood b}), and explained below.
\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= (1-\fp)\prod_{i=1}^N
    \fn^{\theta_i}, \label{eq:likelihood a}\\
    %
    %
    %
    \Pr(\data=1 | \theta) &= 1 - (1-\fp)\prod_{i=1}^N \fn^{\theta_i}. \label{eq:likelihood b}
  %\end{split}
\end{align}
\end{subequations}
A negative pooled test result occurs when all detection paths (both
correct and erroneous) fail. The probability of no false-detection
accounts for the $1-\fp$ term. The probability of no correct detection
is $\fn$ per infected individual. The probability that all such paths
fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}. The probability of a positive result,
presented in \eqref{eq:likelihood b}, is simply the complement.
Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
recalling that $\data \in \{0,1\}$ yields:

>>>>

\section{Introduction}
Reverse transcription polymerase chain reaction (RT-PCR) testing is a
key component in breaking transmission chains and mitigating the
COVID-19 pandemic. As such, the need for large-scale testing has
resulted in the development of techniques to increase the the
throughput of RT-PCR tests \cite{DorfmanYuvalDor, PoolSize30,
  BayesianDorfman, MatrixPooling, LionDorfman}. These techniques are
called \emph{pooling} and originated in the seminal work of Dorfman
pooling \cite{DorfmanOriginal, DorfmanYuvalDor}: Select $N$
individuals and perform a single RT-PCR test on their combined
(\emph{pooled}) samples. If the pooled test yields a positive result
--- test each individual separately. The throughput efficiency of
Dorfman pooling for has been demonstrated empirically
\cite{DorfmanYuvalDor} and its error rates thoroguhly investigated
\cite{Kim, Simplistic1, OptimalDorfmanPool}.

Pooling RT-PCR tests is not perfect, and there is always some nonzero
probability of error. Typically, authors assume the probability of a
true-positive (the test's \emph{sensitivity}) does not depend on the
number of infected samples in the pool, but rather on the existence of
at least one such sample. Thus, the probability of a positive result
in a pooled test is (assumed) identical for a pool with one sample
from an infected individual and (e.g.) five such samples. This
assumption is common in the group testing literature \cite{Kim,
  OptimalDorfmanPool}, as well as in more specific, COVID-19 focused
studies \cite{Simplistic1, Simplistic2}. In this study we refute these
commonly made assumptions and show how they affect estimation of
false-negative rates for Dorfman pooling.

@@

Formally, we consider a pool containing $N$ individuals
$\{1,\dots,N\}$. We denote the true infection states $\theta \in
\{0,1\}^N$, so individual $i$ is infected iff $\theta_i=1$. The RT-PCR
test's sensitivity (true-negative rate) is denoted $\sens$, and the
test's specificity (true-negative rate) is denoted $\spec$. Test results
(data) are denoted $\data$, where $\data=1$ iff the test returned a
positive result.

\subsection{Common assumptions}
Previous studies of group testing strategies assumed that the
false-negative probability does not depend on the number of infected
samples, but merely on the existence of at least one such sample in a
pool \cite{Kim, OptimalDorfmanPool}. Current studies of pooling in the
context of SARS-CoV-2 also employ similar assumptions
\cite{Simplistic1, Simplistic2}. Specifically, these studies assume
that the probability of a negative result is the same for a pool with
a single sample coming from an infected individual and (e.g.)
five. Explicitly:

\begin{align}\label{eq:common}
\Pr(\data=1|\theta) = \left \{
\begin{array}{lr}
        \sens & \exists i\ \text{such that } \theta_i = 1\\
        1-\spec & \text{otherwise}
\end{array} 
\end{align}
We refute the assumptions underlying \eqref{eq:common} in
Section~\ref{subsec:ours} and suggest more plausible ones in
Section~\ref{subsec:ours}.

\subsection{Refuting common assumptions}\label{subsec:refute}
We refute the common assumptions of Section~\ref{subsec:common} with
experimental data summarized in Table \ref{table}, collected from
\cite{Salazar}. There, the authors investigate Dorfman pooling and,
regardless of the pooled test result, follow up and test each pool
member separately. Here, we focus on 128 pools for which at least one
subsequent separate test was positive --- of which 29 pooled tests
were negative and 99 positive. In the data cited in \cite{Salazar}, of
the 29 negative pools, subsequent separate testing yielded a single
positive result in 24. In contrast, of the 99 positive pools, 42
yielded a single positive test upon subsequent separate testing.

The data in Table \ref{table} allows us to test the following null
hypothesis $H_0$: The probability of a pooled false-negative is equal
for pools with one subsequent positively tested member and pools with
two or more such members. We apply Fisher's exact test for the
presence of more than one positive individual in correctly identified
pools. Fisher's test yields an increased odds ratio of 6.4, 95\% CI
(2.2,23.4), with a p-value $\approx 10^{-4}$. Thus we reject $H_0$,
refuting the independence of a pooled result from one or more
individuals infected within the pool, as assumed in \cite{Simplistic1,
  Simplistic2, OptimalDorfmanPool, Kim}

@@

The probability of a negative pooled test result is presented in
\eqref{eq:likelihood a} (along with its complement
\eqref{eq:likelihood b}), and explained below.
\begin{subequations}
\begin{align}%\label{eq:single likelihood}
  %\begin{split}
    \Pr(\data=0 | \theta) &= (1-\fp)\prod_{i=1}^N
    \fn^{\theta_i}, \label{eq:likelihood a}\\
    %
    %
    %
    \Pr(\data=1 | \theta) &= 1 - (1-\fp)\prod_{i=1}^N \fn^{\theta_i}. \label{eq:likelihood b}
  %\end{split}
\end{align}
\end{subequations}
A negative pooled test result occurs when all detection paths (both
correct and erroneous) fail. The probability of no false-detection
accounts for the $1-\fp$ term. The probability of no correct detection
is $\fn$ per infected individual. The probability that all such paths
fail is the product of the above mentioned terms, displayed in
\eqref{eq:likelihood a}. The probability of a positive result,
presented in \eqref{eq:likelihood b}, is simply the complement.
Combining \eqref{eq:likelihood a} and \eqref{eq:likelihood b}, and
recalling that $\data \in \{0,1\}$ yields:
